\documentclass[../main.tex]{subfiles}
\graphicspath{{\subfix{../images/}}}
\begin{document}

First and foremost, I would like to thank my supervisor, Mariami Gachechiladze, for accepting me into your newly founded group despite my complete lack of knowledge about quantum computing. Never in my academic life have I received such hands-on supervision. Despite your busy schedule, you always found time to lend a helping hand when needed. It was incredibly valuable to witness firsthand your passion and commitment in building a research group from scratch.

I also want to thank Philipp Harbach, Mathias Winkel, and all members of the Group Digital Innovation, who granted me complete freedom to pursue my studies and provided opportunities to participate in exciting research collaborations.

I want to thank my academic collaborators, Matthias Heller and Federico Tiblias. I really appreciated having collaborators so nearby, allowing us to discuss regularly in person.

I also want to thank the members of the QUTAC Materials Science Group, who warmly welcomed me into the working group. You not only allowed me to learn about the very frontiers of quantum computing for chemistry, but also showed me, by example, how to build a community for collaborative research. The opportunity to serve as the administrative lead for a period provided me with valuable growth experiences.

Thank you to group members Jan Nöller, Prabhav Jain, Robin Krebs, Lucas Vieira, and Minh Duc Tran for the unforgettable times at conferences and entertaining lunches at h\_da Mensa. You are wonderful people, and I hope to keep in touch with you.

I want to thank everyone who proofread my thesis: Prabhav Jain, Maximilian Scheurer, Federico Tiblias, and Lucas Vieira.

Also, thank you to those who provided much-needed emotional support during my many mental breakdowns: Elif Yildirim and Prabhav Jain.

I would like to especially thank my partner Simeon Scheib, who has listened patiently to my whinings over so many years. You have always been my greatest supporter.

Thank you to Momo the cat. It is the highest honor to have you on the cover page.

Last but not least, I would like to extend my gratitude to my family, who ultimately shaped me into who I am today. This thesis is dedicated to my sister, my mother, and to the memory of my father.