\documentclass[../main.tex]{subfiles}
\graphicspath{{\subfix{../images/}}}
\begin{document}

\section{Preface}

Quantum computing is, by its very name, an interdisciplinary field that attracts physicists, engineers, computer scientists, mathematicians, as well as recently chemists, biologists, and many other researchers from quantitative sciences. This interdisciplinary endeavor is fueled by the joint aim to leverage quantum physics to solve problems that are difficult to address with conventional classical computing.
An intuition propagated by Richard Feynman in 1981, at the first conference on "Physics of Computation," was that a quantum computer could simulate quantum systems more efficiently than a classical computer. At the same conference \cite{history_quantum}, Paul Benioff presented his work on a Turing machine operating under quantum mechanics. Only a decade later, Deutsch and Jozsa established the first complexity-theoretic proof for a problem instance, for which the quantum algorithm is efficient while no efficient deterministic classical solution exists.

In 1994, Peter Shor made a groundbreaking discovery, demonstrating that quantum computers can efficiently solve a problem in NP. Shor's algorithm efficiently implements prime factorization. Since this particular computational hardness is ubiquitous in cryptography, this discovery was not only exciting regarding the complexity separation between classical and quantum computers, but also came with profound practical and societal implications due to its potential to jeopardize any prevalent encryption scheme.
Hence, these breakthroughs justified increased funding for experimental endeavors, which resulted within three decades in a Nobel Prize and numerous platforms based on different physical phenomena to implement quantum computers. Yet another inflection point was marked by Google's publication in 2019 \cite{arute2019quantum}, which claimed the first experimental demonstration of a computational task solved by a quantum computer that could not be conducted classically. While that particular task, random circuit sampling, does not pose any relevance to practical use cases, this strong claim triggered interest among industry leaders seeking to comprehend how quantum algorithmic research might change their businesses. It also increased funding for academia and attracted substantial private investments for startups providing hardware, software, and entire ecosystems around quantum computing, to such an extent that many people in the field characterize the current state as a hype \cite{sarma2022quantum,hoefler2023disentangling}.

In this regard, the author's endeavors align well with this trajectory. Dispatched from a pharmaceutical company, the aim was to comprehend this field better, to distinguish the hype from actual claims, to obtain sufficient foundations such that both the past and the state-of-the-art of quantum computing research can be understood, and to grow into someone who can track the advances to come, while exploring use cases of quantum computing within the application realm of quantum chemistry.

This thesis comprises five chapters, including this chapter, which provides basic introductions to quantum mechanics, quantum computing, and a class of algorithms known as variational quantum eigensolvers.

The research results accumulated over the past three and a half years are presented as standalone works. First, we present an alternative formulation of the variational quantum eigensolver using a different paradigm of quantum computing, the measurement-based quantum computer, in Chapter~\ref{chap:mbvqe}. The original proposal of this measurement-based eigensolver, Ref.~\cite{ferguson2021measurement}, came with an exponential sampling overhead, which we mitigate by invoking known results from deterministic strategies in measurement-based quantum computing.
Subsequently, results on applications of variational quantum eigensolvers in the chemistry domain are presented in Chapter~\ref{chap:chem}. This work was conducted within the collaborative efforts of the Quantum Technology and Application consortium and studies the efficacy of a quantum-classical hybrid algorithm, the quantum-enhanced auxiliary-field quantum Monte Carlo, for estimating ground state energies of molecules, as well as the overall scaling to assess applicability beyond toy system settings.
We then discuss the application of variational quantum eigensolvers in the setting of classical machine learning tasks, where we test a data-encoding method inspired by encodings inherently given in quantum spin systems in Chapter~\ref{chap:qnlp}. 
Last but not least, we gain abstract-level intuition on quantum advantage by examining memory advantages in quantum computing that are present in so-called promise problems for finite state automata, which have been studied since the early days of quantum computational complexity theory in the 1990s and were recently proposed for certification of quantum models in Chapter~\ref{chap:automaton}. We propose several novel classes of promise problems with quantum-classical complexity separation.

Each chapter includes its own introduction and theoretical background. This work thus encompasses a wide range of theoretical branches in quantum computing, including measurement-based quantum computing, applications in quantum chemistry and machine learning, and decision problems with quantum advantages. 

\section{Quantum Computing 101 \& Notations}\label{sec:quantum101}

Quantum computers perform arithmetic operations using quantum systems, typically quantum two-level systems, as opposed to classical two-level systems characterized by binary integers, the elementary information units of what the quantum computing community refers to as classical computers.

In the following sections, we summarize the basic concepts and notations of quantum computing to make this thesis as self-contained as possible. We note that numerous excellent textbooks exist for an introduction to quantum computing and quantum information science \cite{nielsen2010quantum, wong2022introduction, watrous2018theory}, as well as online resources provided by commercial quantum computing companies \cite{2025learn, 2025quantum}.

The mathematical formulations required to treat quantum systems follow from the four \textit{postulates of quantum mechanics}. We consider exclusively closed and finite-dimensional quantum systems and outline these postulates with accompanying comments:
\begin{enumerate}[align=left]
        \item[Postulate 1] \textit{State Space}: Any isolated physical system is associated with a Hilbert space $\mathcal{H}$, i.e., a complex vector space $\mathbb{C}^n$ equipped with an inner product $\langle \cdot | \cdot \rangle : \mathbb{C}^n \times \mathbb{C}^n \mapsto \mathbb{C}$. \textit{Dirac notation} is employed to concisely express high-dimensional vectors $\ket{\phi} \in \mathbb{C}^n$ and their conjugate transpose $\bra{\phi}$. The physical system is fully described by the \textit{state vector}, e.g., $\ket{\phi}$, a unit vector in the respective Hilbert space with $\braket{\phi}{\phi} = 1$.
        
    \item[Postulate 2] \textit{Evolution}: The time evolution of an isolated physical system is           governed by the Schrödinger equation
        \begin{equation}
            i \hbar \frac{\dd \ket{\phi}}{\dd t} = H \ket{\phi},
        \end{equation} where $H$ is the \textit{Hamiltonian}, a Hermitian operator, i.e., $H = (H^{*})^T = H^\dagger$, that corresponds to the total energy of the system. Mathematically, the evolution operator is given by a \textit{unitary} operator, i.e., $U \ket{\phi} = \widetilde{\ket{\phi}}$ with $UU^\dagger = U^\dagger U = \mathds{1}$, mapping a state vector $\ket{\phi}$ to another valid state vector $\widetilde{\ket{\phi}}$. In other words, $U$ implements a norm-preserving rotation in the Hilbert space.
        
    \item[Postulate 3] \textit{Measurement}: Measurements on a quantum system are described by a set of measurement operators ${M_a }$, where the index $a$ denotes a possible outcome. Given a system $\ket{\phi}$, the probability that $a$ is measured is given by 
    \begin{equation}
        p(a) = \bra{\phi} M_a^\dagger  M_a \ket{\phi}
    \end{equation} and the post-measurement state is 
    \begin{equation}
        \frac{M_a \ket{\phi}}{\sqrt{\bra{\phi} M^\dagger_a M_a \ket{\phi} }}.
    \end{equation} The measurement operators obey the completeness relation
    \begin{equation}
        \sum_a M_a^\dagger M_a = \mathds{1},
    \end{equation} which follows from the natural requirement that the probabilities sum to $1$. If the measurement operators are projective operators, the measurement is called a \ac{PVM}. This is the typical case in quantum computing.
    
    \item[Postulate 4] \textit{Composition}: The state space of a composite system can be expressed as the \textit{tensor product} of its constituent systems, e.g., $\mathcal{H}_{12} = \mathcal{H}_{1} \otimes \mathcal{H}_{2}$. The same holds for composite state vectors 
    $\ket{\phi} = \ket{\phi_1} \otimes \ket{\phi_2} \otimes \cdots \otimes \ket{\phi_m}$. If $\ket{\phi_i} \in \mathbb{C}^{n}$, then $\ket{\phi} \in \mathbb{C}^{n^m}$.
        
\end{enumerate}

We introduce some common terminology.  A \textit{qubit} is a quantum system with dimension two, with the general state given by $\alpha \ket{0} + \beta \ket{1}$ where $|\alpha|^2 + |\beta|^2 = 1$, and $\ket{0} = (1 \quad 0)^T$ and $\ket{1} = (0 \quad 1)^T$. These one-hot vectors are referred to as the \textit{computational basis} or the Pauli $Z$-basis, and this notion extends to composite systems, or multi-qubit systems, through the tensor product. It is convenient to work with \textit{Pauli matrices}
\begin{equation}
    X = \begin{pmatrix}
        0 & 1 \\
        1 & 0 
    \end{pmatrix}, \quad
    Z =
    \begin{pmatrix}
        1 & 0 \\
        0 & -1 
    \end{pmatrix}, \quad
    Y = iXZ = 
    \begin{pmatrix}
        0 & -i \\
        i & 0
    \end{pmatrix},
\end{equation} which are unitary and form, together with the identity matrix $\mathds{1}$, a basis for $2 \times 2$ Hermitian matrices. \textit{Pauli words} or \textit{Pauli strings} are multi-qubit operators consisting of Pauli matrices composed through tensor products $\bigotimes_n P_i$, where $P_i \in \{X,Y,Z,\mathds{1} \}$. These are exactly the kinds of matrices that are relevant as physical operators or observables. \textit{Superposition} describes a quantum state that exists as a linear combination of basis states simultaneously; e.g., for non-zero $\alpha$ and $\beta$, $\alpha \ket{0} + \beta \ket{1}$ is a superposition of $\ket{0}$ and $\ket{1}$. \textit{Entanglement} occurs in composite systems when the state vector cannot be expressed as a product state $\ket{\phi}_{AB} \neq \ket{\phi}_A \otimes \ket{\phi}_B$. An example of such entangled states are Bell states, for example $\frac{1}{\sqrt{2}} (\ket{00} + \ket{11})$, where $\ket{ij} = \ket{i} \otimes \ket{j}$, which together with the Bell inequality \cite{bell1964einstein} proved the incompatibility of quantum theories with local hidden-variable theories given the statistical correlations such states produce through measurements. Entanglement serves as a communication resource \cite{ekert1991quantum}, a computation resource in measurement-based quantum computing \cite{raussendorf2003measurement}, and is considered a necessary condition for many instances of quantum advantage \cite{vidal2003efficient}.

However, it is only a necessary condition, as there are highly entangled quantum states that can still be simulated efficiently by classical means. One example of such states are the so-called \textit{Stabilizer states}. These states are precisely those that can be completely characterized by their Pauli string stabilizers, i.e., the $+1$ eigenstates corresponding to certain Pauli operators. Such states can be mapped onto each other via \textit{Clifford circuits}, which can be generated by a finite set of gates, e.g., the \textit{Hadamard gate}

\begin{equation}
    H = \frac{1}{\sqrt{2}}
    \begin{pmatrix}
        1 & 1 \\
        1 & -1
    \end{pmatrix},
\end{equation} the $S$\textit{-gate}

\begin{equation}
    S = \sqrt{Z} =
    \begin{pmatrix}
        1 & 0 \\
        0 & i
    \end{pmatrix},
\end{equation} and the two-qubit controlled-NOT ($\mathrm{CNOT}$)
\begin{equation}
    \mathrm{CNOT} = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        0 & 0 & 0 & 1 \\
        0 & 0 & 1 & 0 \\
    \end{pmatrix}.
\end{equation} Classical simulability (i.e., the feasibility of efficiently simulating certain evolutions on a classical computer) of these states under Clifford gates was proven under the \textit{Gottesman-Knill theorem} \cite{aaronson2004improved} as a constructive algorithm that essentially keeps track of how these elementary gates map one Pauli string to another. For instance, we have $HXH^\dagger = Z$ and $HZH^\dagger = X$, $SXS^\dagger = Y$ and $SZS^\dagger = Z$, and $\mathrm{CNOT} (X \otimes I) \mathrm{CNOT}^\dagger = X \otimes X$, $\mathrm{CNOT} (I \otimes X) \mathrm{CNOT}^\dagger = I \otimes X$, $\mathrm{CNOT} (Z \otimes I) \mathrm{CNOT}^\dagger = Z \otimes I$ and $\mathrm{CNOT} (I \otimes Z) \mathrm{CNOT}^\dagger = Z \otimes Z$.  Another class of classically simulable states is the \textit{free Gaussian states}, which we will introduce later in \ref{subsub:matchgate}. Clifford gates are not \textit{universal}, i.e., they cannot evolve states to any arbitrary state, since the image is restricted to stabilizer states. However, augmenting any non-Clifford circuit to the Clifford gates, such as the $T$-gate
\begin{equation}
    T = \sqrt{S} = \begin{pmatrix}
        1 & 0 \\
        0 & e^{i \frac{\pi}{4}}
    \end{pmatrix},
\end{equation} renders the set to be the approximate generator of the set of all unitaries. The discovery that a finite set of generators can approximate any unitary operation represents a significant step toward physically realizing universal quantum computers.
 However, the ability to approximate universal gates does not necessarily guarantee efficiency. From the Solovay-Kitaev theorem, we know that approximating an arbitrary $n$-qubit unitary $U$ up to distance $\epsilon$ takes $O(n^2 4^n \log^c(n^24^2/\epsilon))$ for some constant $c$ \cite{nielsen2010quantum}. 
 
 These generating sets are not unique. Another popular choice is the combination of general single-qubit rotations with $\mathrm{CNOT}$ $\{R_x(\theta_x), R_y(\theta_y), R_z(\theta_z),  \mathrm{CNOT}\}$, where
\begin{align}
        R_x(\theta) &= e^{-i \frac{\theta}{2} X } = 
    \begin{pmatrix}
        \cos{\frac{\theta}{2}} & -i \sin{\frac{\theta}{2}} \\
        -i \sin{\frac{\theta}{2}} & \cos{\frac{\theta}{2}}
    \end{pmatrix}, \\
        R_y(\theta) &= e^{-i \frac{\theta}{2} Y } = 
    \begin{pmatrix}
        \cos{\frac{\theta}{2}} & - \sin{\frac{\theta}{2}} \\
        - \sin{\frac{\theta}{2}} & \cos{\frac{\theta}{2}}
    \end{pmatrix}, \\
        R_z(\theta) &= e^{-i \frac{\theta}{2} Z } = 
    \begin{pmatrix}
        e^{-i \frac{\theta}{2} } & 0 \\
        0 & e^{i \frac{\theta}{2} }
    \end{pmatrix}.
\end{align} The $\mathrm{CNOT}$ in combination with any choice of two among these three rotation gates is sufficient for universal quantum computing.

\subsection{Measurements in quantum computing}\label{sec:measure}
A \textit{quantum algorithm} consists largely of three steps: 1. the initial state preparation, including some form of data loading; 2. processing through gate sequence applications; and 3. the measurement. Details on initial state preparation and concrete gate sequences will be introduced in this thesis where appropriate (e.g., Chap.~\ref{subsub:quantumquantumchem}). Here, we focus on the measurement process.

Some of the key distinctions between classical and quantum computing can be traced to the effects of measurement. While read-out of classical bits is ``trivial'' since they are represented completely by a single observable (in quantum computing terminology, an eigenstate of the $Z$ operator), a qubit is generally characterized by a set of non-commuting operators. As mentioned in Postulate 3, measurements alter the measured state, preventing inference of additional information from a single copy. The uncertainty introduced through measurements of incompatible (i.e., non-commuting) observables is known as Heisenberg's uncertainty relation. It has also been shown that no unitary operation can exist on a two-qubit system that can copy an unknown quantum state from one qubit to another. This limitation is known as the \textit{no-cloning theorem}.  We will now discuss some strategies for extracting data from a quantum computer through measurements, nonetheless.

A measurement type commonly supported by quantum hardware is the projective measurement onto the computational basis. This offers sufficient functionality to measure in any basis since we can append unitaries to the circuit that rotate the state to the desired measurement basis. For example, if we want to measure in the $X$-basis, we apply $H$ since $HZH^\dagger = X$, or $SH$ if we want to measure in the $Y$-basis due to $SHZ(SH)^\dagger = Y$. 
Expectation values are estimated via the sample mean with some post-processing. E.g., if we are interested in estimating $\bra{\psi}Z\ket{\psi}$, we measure in the $Z$ basis and associate a $+1$ whenever we project onto $\ket{0}$ and $-1$ when we project onto $\ket{1}$, corresponding to their respective eigenvalues. In other words, we obtain a binomial distribution, and the error of the expectation value (or the mean) scales according to the central limit theorem as $\propto 1/\sqrt{N}$, where $N$ is the number of samples. 
The same principle holds for Pauli words and multi-qubit systems. Since Pauli matrices are both unitary and Hermitian, they too have eigenvalues of either $+1$ or $-1$. For a given Pauli word, the appropriate basis rotations are applied to each qubit, and the output is computed as the parity of the resulting binary string.

\begin{figure}[h!]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \begin{quantikz}
            \ket{0} & \gate{H} & \ctrl{1} & \gate{H} & \meter{} \\
            \ket{\psi} && \gate{U} & &
        \end{quantikz}
        \caption{circuit for estimating $\mathrm{Re}(\bra{\psi}U\ket{\psi})$}
    \end{subfigure}\hfill
    \begin{subfigure}{0.45\textwidth}
        \centering
        \begin{quantikz}
            \ket{0} & \gate{H} & \gate{S^\dagger} & \ctrl{1} & \gate{H} & \meter{} \\
            \ket{\psi} &&& \gate{U} & &
        \end{quantikz}
        \caption{circuit for estimating $\mathrm{Im}(\bra{\psi}U\ket{\psi})$}
    \end{subfigure}
    \caption{Circuits for the Hadamard test}\label{fig:hadamard}
\end{figure}


Another routine for measuring expectation values of a unitary operator $\bra{\psi} U \ket{\psi}$ is the \textit{Hadamard test} \cite{nielsen2010quantum, tilly2022variational}. Here, in addition to the qubits implementing the state of interest $\ket{\psi}$, an auxiliary qubit is used, which is prepared in the $\ket{+}$ state. By applying a controlled version of the unitary $U$ with the auxiliary qubit as the control, the expectation value is encoded in the relative phase of the auxiliary qubit, such that measuring the auxiliary qubit suffices to obtain the expectation value. The sampling complexity scales as $O(1/\epsilon^2)$, the typical statistical shot-noise scaling, where $\epsilon$ is the desired precision, and is independent of the system size. Note, however, that this method estimates only the real part of the expectation value. To estimate the imaginary part, the initial state of the auxiliary qubit must be modified --- the circuits implementing this routine are shown in Fig.~\ref{fig:hadamard}.

While in algorithms such as Shor's algorithm \cite{shor1994algorithms}, the output is formulated as a computational basis state, in more physically motivated problems where a quantum computer is used to prepare a specific quantum state of interest, the aim is often to determine the physical properties of that state. These properties are given in the form of Hermitian operators, which are typically not unitaries. Thus, these observables need to undergo preprocessing to unitarize them or to express them, for instance, as an \ac{LCU}. The most convenient approach is to express operators as sums of Pauli words, since physical operators are canonically given in this form. More advanced decomposition methods are summarized in \cite{loaiza2023reducing}. These methods require higher preprocessing costs but offer favorable operator properties for subsequent quantum applications. Given an LCU representation of an operator, we can sequentially measure each unitary component using the methods described above and weight the expectation value with the respective coefficient in the LCU.

Probably the most powerful and efficient method for measuring expectation values from a quantum computer is \ac{QPE} \cite{kitaev1995quantum, nielsen2010quantum}. This is a one-shot algorithm that computes eigenvalues when given access to a unitary operator $U$ and its eigenstate of interest $\ket{\psi}$, often the ground state of a given Hamiltonian $H$. Since $H$ is not unitary, one instead considers the operator $U = e^{-2\pi iH}$, which preserves eigenstates and transforms the eigenvalue to a phase: $U|\psi\rangle = e^{2\pi i\theta}|\psi\rangle$, where $\theta$ is the quantity we want to estimate. In \ac{QPE}, the value of $\theta$ is computed and encoded as an $n$-bit binary representation using $n$ auxiliary qubits and a register containing the prepared state $|\psi\rangle$. The algorithm employs the inverse quantum Fourier transform (QFT), which is the algorithmic primitive used in other quantum algorithms with exponential speedups, such as Shor's algorithm and discrete logarithm algorithms. The \ac{QFT} as a blackbox function is
\begin{equation}
    \ket{x} = \sum_{i=0}^{2^n - 1} x_i \ket{i} \xrightarrow{QFT} \ket{y} = \sum_{i=0}^{2^n - 1} y_i \ket{i}, \quad y_k = \frac{1}{2^n} \sum_{j=0}^{2^n - 1} x_j e^{\frac{2 \pi i k j}{2^n}}.
\end{equation} 
Note that if $\ket{x} = \ket{k}$, a computational basis state, $ \ket{x} \xrightarrow{QFT} \frac{1}{2^n} \sum_{\ell = 0} e^{2 \pi i \tilde{k} \ell} \ket{\ell}$, where $\tilde{k} = k/2^{n}$ and is a binary fraction $\tilde{k} = \sum_{i=0} k_i 2^{n-1-i}$. For the phase estimation problem, we want to prepare the state $\frac{1}{\sqrt{2^n}} \sum_\ell e^{2 \pi i \theta \ell} \ket{\ell}$ so that we can map back to $\ket{\theta}$ with the inverse \ac{QFT}, then measure in the computational basis to obtain the eigenvalue deterministically with precision depending solely on the number of auxiliary qubits. We observe that  $U^k \ket{\psi} = e^{2 \pi i \theta k} \ket{\psi}$. Thus, applying binary powers of $U$ controlled by the superposition of all computational states  $\frac{1}{\sqrt{2^n}} \sum_k \ket{k}$ gives exactly the desired state. This brief description is illustrated in Fig.~\ref{fig:qpecircuit}.

Since this routine offers the possibility of estimating ground state energies without solving the diagonalization problem of an intractably large matrix, it represents a very promising candidate for practical quantum advantage. However, challenges arise when attempting to actually implement this algorithm. 
First, the preparation of $\ket{\psi}$ as the exact ground state is challenging. In realistic implementations, $\ket{\psi}$ would be rather a superposition of multiple eigenstates. This makes the output of \ac{QPE} also a superposition of different bit strings, rendering it a probabilistic protocol where the success probability depends mostly on the overlap between the initial state and the desired eigenstate. Another challenge arises from implementing the controlled unitaries. Typically, $H$ is a sum of many non-commuting terms, making the implementation of $e^{ixH}$ possibly intractable even on quantum computers due to the high circuit depths required. High circuit depth also means that hardware errors will almost certainly corrupt the output, as errors accumulate with each gate operation. Strategies for implementing Hamiltonian evolution operators, including Trotter decomposition, will be discussed later in Chap.~\ref{subsub:quantumquantumchem}. Many follow-up works have aimed to reduce the required hardware resources by increasing the sampling complexity for more favorable qubit counts \cite{kitaev1995quantum, aspuru2005simulated} or circuit depth \cite{wan2021a, dutkiewicz2024error}. Due to such issues, \ac{QPE} is considered a quantum algorithm accessible only on fault-tolerant quantum computers, i.e., those free of hardware noise, which are not yet an experimental reality.

\begin{figure}
        \centering
        \begin{quantikz}
            \ket{0} & \gate{H} &&&  \  \ldots \  & \ctrl{4}& \gate[4]{\mathrm{QFT}^{-1}} & \meter{} \\
            \vdots \setwiretype{n}&&&&\setwiretype{n}  \  \vdots \  &&&\setwiretype{n} \vdots \\
            \ket{0} & \gate{H} & & \ctrl{2} & \  \ldots \  &&  & \meter{} \\
            \ket{0} & \gate{H} & \ctrl{1} && \  \ldots \  &&  & \meter{}  \\
            \ket{\psi} && \gate{U^{2^0}} & \gate{U^{2^1}} & \  \ldots \  & \gate{U^{2^{n-1}}} & &
        \end{quantikz}
        \caption{Quantum circuit implementing the ``textbook'' QPE \cite{kitaev1995quantum, nielsen2010quantum} to compute the phase $\theta$ given a unitary $U$ and eigenstate $\ket{\psi}$ sucht that $U \ket{\psi} = e^{2 \pi i \theta} \ket{\psi}$. $n$ auxiliary qubits and controlled applications of powers of $U$ are used to encode the Fourier coefficients into the phase of the auxiliary qubits. Then, the inverse \ac{QFT} is applied to resolve $\theta$ in binary representation in the output register. }\label{fig:qpecircuit}
\end{figure}

There has been recent innovation in \ac{NISQ}-compatible measurement schemes, called \textit{classical shadows} \cite{huang2020predicting}, which is based on building a bias-free estimator from measurement data that can be used as a post-processing routine to estimate many observables classically and efficiently, making it attractive for quantum-classical hybrid schemes. We will discuss this topic in detail when this scheme is utilized in such contexts (see Sec.~\ref{sub:shadowsandstats}).

\subsection{Path to Fault-tolerant QC}

As a final section of this brief summary of quantum computing, we mention that inevitable hardware and environmental noise affecting qubits and quantum gates can be overcome through error-correcting codes. We note that classical bits, implemented at the elemental level through transistors etched on silicon wafers at atomic scales, are also prone to bit-flip errors and similar issues, but are successfully treated through error-correcting codes typically based on redundancy of stored information. In the simplest examples, many copies of the same physical bit are created to form one logical bit. While the no-cloning theorem prohibits \textit{to copy} the same strategy to build fault-tolerant quantum computers, efforts have been made with success to develop quantum error-correcting codes theoretically since the dawn of quantum computing \cite{shor1995scheme, steane1996error, gottesman1997stabilizer}. Since this field is not the focus of this thesis, we reference  \cite{roffe2019quantum, terhal2013quantum} for details. Redundancy in quantum error correction is implemented via stabilizer states, since with these, we know how to perform measurements in the eigenbasis without disturbing the underlying logical state. The parity of such stabilizer measurements reveals the existence of certain error types, which can then be compensated via gates conditioned on these so-called syndrome measurement outcomes. The most popular codes are probably topological or surface codes  \cite{fowler2012surface}, based on stabilizer states consisting of local \textit{star} and \textit{plaquette operators} named after the local structure of the respective operators on a two-dimensional square grid. If the error probability per gate is below a certain threshold, it has been shown that errors can be arbitrarily reduced in the limit of large code sizes. This is known as the threshold theorem \cite{shor1996fault}, and the concrete thresholds depend on the error-correcting code. For the surface code, this critical threshold has been estimated through numerical studies accounting for realistic error scenarios to be approximately $0.1094 \pm 0.00020$ \cite{dennis2001topological}.

First demonstrations of error suppression using surface codes have been shown by multiple commercial hardware providers using different technology platforms: superconducting qubits, neutral atoms, and photonic quantum computing  \cite{acharya2022suppressing, bluvstein2023logical, walshe2025linear}. With such advancements in hardware development, practical quantum advantage appears imminent. However, the field has not remained idle and has spent more than a decade contemplating whether practical quantum advantage is achievable even on \ac{NISQ} computers, a limited quantum device with about 50-100 qubits that can implement circuits with a depth of below 100 \cite{preskill2018quantum}, which we will introduce in the next chapter.


\section{Noisy intermediate scale quantum computing}

While progress on quantum hardware is impressive, with recent demonstrations providing evidence of error correction using small instances of error-correcting codes \cite{acharya2022suppressing, bluvstein2023logical, walshe2025linear}, currently existing machines are categorized as \ac{NISQ}-devices \cite{preskill2018quantum}. There appears to be a ``cat-and-mouse game'' between those who claim experimental demonstrations of quantum advantage and those who attempt to disprove these claims by devising classical methods that compute the same results with reasonable classical resources. For instance, Google's quantum ``supremacy''\footnote{This terminology is used as a direct citation; in general, it is advised to replace it with a more historically neutral term as appealed here \cite{palacios2019instead}} claim through random circuit sampling  \cite{arute2019quantum} was countered by Pan et al.~\cite{pan2022solving}, who claimed classical simulability.  Similarly, IBM's work claiming utility in quantum simulation of the transverse-field Ising model \cite{kim2023evidence}, immediately prompted counter-papers demonstrating classical simulation through DMRG-based methods \cite{tindall2024efficient} and near-Clifford approaches \cite{begusic2023fast}. However, given that disproving quantum advantage claims now requires significant intellectual effort compared to earlier demonstrations such as factoring the number 15 \cite{vandersypen2001experimental}, it is apparent that the computational power of noisy devices is becoming worthy of consideration for practical quantum computing applications. Yet hardware noise remains too overwhelming to implement quantum algorithms with theoretically proven complexity advantages. A line of research that emerged from the desire to make use of noise-constrained computing is the class of \ac{VQA}, or specifically \ac{VQE}. In the following section, we will examine what this class of algorithms does, their promises, and their caveats.


\subsection{Variational Quantum Eigensolver}\label{sec:vqe}

The variational method, also known as the Rayleigh–Ritz method, is an approximate technique for solving eigenvalue problems. In quantum physics, this method is applied when one seeks the ground state and corresponding energy of a known Hamiltonian $H$, but an analytic solution is deemed infeasible \cite{zettili2009quantum}. Such situations commonly occur in quantum chemistry, a discipline focused on predicting molecular and material properties from quantum-mechanical first principles, which is inherently hard because the interaction space typically grows exponentially with the number of electrons involved. A concise overview of how these problems are mathematically modeled will be provided later in Chap.~\ref{sec:introquantumchem}. 

In the Rayleigh–Ritz method, a parameterized \textit{trial wave function} $\ket{\psi(\bm{\theta})}$ is introduced, where $\bm{\theta} = (\theta_0, \dots, \theta_n)$. Within the quantum-computing community, this trial wave function is commonly called \textit{ansatz}. The expectation value is then calculated as 
\begin{equation}
    E (\bm{\theta}) = \expval{H}{\psi{(\bm{\theta})}}\geq E_{gs} \;,
\end{equation} where this energy estimate is always bounded from below by the exact ground-state energy $E_{gs}$. The reason is that $\ket{\psi(\bm{\theta})}$ remains a valid state vector for any parameter choice, and $E_{gs}$ represents the minimal eigenvalue. The aim is to find parameter values such that $E (\bm{\theta})$ is minimized. This minimization typically involves three steps:
\begin{enumerate}
    \item Initialization 
    \item Optimization (using numerical methods such as gradient-based, gradient-free, or analytical techniques)
    \item Evaluation 
\end{enumerate}
Once the estimated energy is satisfactory, the procedure is stopped. Commonly, ansätze used in the literature, e.g., Hartree-Fock \cite{slater1951simplification}, or \ac{DMRG}  \cite{schollwock2005density}, include simplifications that make the process numerically feasible but restrict the method to heuristic use. The idea behind the Variational Quantum Eigensolver (VQE) \cite{peruzzo2014variational,kandala2017hardware,tilly2022variational,mcclean2016theory}, which remains heuristic, is to implement classically intractable ansätze on a quantum computer while carrying out optimization steps classically to reduce quantum hardware resources. Hence, VQEs are referred to as quantum–classical hybrid algorithms. 

\subsubsection{Ansätze in VQE}

We now briefly outline the steps involved in implementing a \ac{VQE}.
The first step is to decide which ansatz to use. In VQEs, these ansätze are parameterized unitary operations $U(\bm{\theta})$, and different kinds have been proposed to address various priorities. 
To mention some, there is the \textit{Hardware-efficient} ansatz utilizing rotation axes and entangling gates native to the specific hardware platform without incurring compilation overhead. Entangling operations are restricted to qubits with direct-neighbor interactions to avoid SWAP gates, and local rotation layers alternate with entangling layers. As its name suggests, this ansatz optimizes hardware implementation efficiency relative to algorithmic circuit depth, making it particularly popular in early VQE hardware demonstrations \cite{kandala2017hardware}. While it can, in principle, approximate any quantum state, meaning no domain-specific information is incorporated, it suffers from barren plateaus \cite{holmes2022connecting}, i.e., vanishing gradients rendering the optimization processes intractable.
The \textit{unitary coupled-cluster} ansatz was the first proposed within the quantum–classical hybrid framework. We will discuss its details in Section~\ref{subsub:quantumquantumchem}, but note here that coupled-cluster theory is considered the ``gold standard''  in quantum chemistry for benchmarking groundstate energies beyond Hartree–Fock calculations \cite{bartlett2007coupled}. It employs an exponential ansatz that cannot be implemented classically in a tractable manner. However, its unitarized version can be efficiently realized on quantum devices \cite{peruzzo2014variational, mcclean2016theory}. Nevertheless, such ansätze typically scale as $\mathcal{O}(N^4)$ in circuit depth, where $N$ is the number of qubits, or spin orbitals, making them impractical to implement on NISQ-devices for larger systems that are not classically treatable.

Two popular methods combine the advantages of hardware-efficient and problem-aware unitary coupled-cluster approaches: The \textit{quantum number preserving} (QNP) ansatz \cite{anselmetti2021local} identifies minimal local gate blocks (restricted to nearest-neighbor interactions) that preserve essential quantum numbers during ground-state calculations, such as particle number and spin symmetries. These blocks enable the construction of ansätze that respect symmetry constraints and device limitations. 
Further, \textit{ADAPT}-VQE \cite{grimsley2019adaptive} offers an alternative approach to reducing the complexity of unitary coupled-cluster ansätze. Rather than including all terms from the unitary coupled-cluster ansatz, this method iteratively grows the ansatz by appending only those terms that exhibit the largest gradients with respect to the energy expectation value. This adaptive method systematically builds up circuit depth while prioritizing the most impactful terms for energy minimization.

\subsubsection{Gradient estimation in VQE}

How are gradients extracted within VQE? VQE typically addresses cases where analytical computation of the optimum $\frac{\partial E(\bm{\theta})}{\partial \theta_i} = 0$ is prohibitive, necessitating the use of numerical optimizers. In prototypical studies involving test systems small enough to permit classical simulation of the entire workflow, researchers can employ either gradient-free methods such as COBYLA \cite{ragonneau2023pdfo:}, or gradient-based approaches such as L-BFGS \cite{liu1989on}, and popular machine-learning optimizer ADAM \cite{kingma2014adam}. Gradients, and higher-order partial derivatives when necessary, can be estimated through finite difference approximation or computed using automatic differentiation \cite{paszke2017automatic}. The latter approach has been recently incorporated into quantum computing software libraries such as PennyLane \cite{bergholm2018pennylane}, enabling convenient computation for VQE.

For future settings where quantum ansätze are anticipated to have parameter counts beyond what classical gradient estimation schemes can manage, analytical gradients can be estimated using the quantum computer itself. This method, called \textit{parameter shift rules} \cite{mitarai2018quantum, schuld2018evaluating}, exploits the fact that gradients of expectation values from unitary ansätze can be expressed exactly as differences of the same operators evaluated with shifted parameters (typically $\theta \pm \pi/2$, where $\theta$ is the current parameter value). While this approach provides exact gradients without approximation error, it requires additional circuit evaluations proportional to the number of parameters.

\subsubsection{Evaluation and measurements in VQE}

The evaluation of expectation values is crucial for the practicality of the entire workflow. As mentioned in Sec.~\ref{sec:measure}, \ac{QPE} cannot be employed on NISQ hardware, requiring the use of sampling-based schemes. The variance of each measured term scales with the shot noise limit $\mathcal{O}(1/\epsilon^2)$. The number of terms depends on the cost function-defining operator; for the electronic-structure Hamiltonian, the number of Pauli strings scales as $\mathcal{O}(N^4)$, where $N$ is the number of qubits. For computational chemistry, a typical target precision is the \textit{chemical accuracy} $\epsilon = 1.6 \mathrm{mH}$. This value is said to match the accuracy of experimental measurements in chemistry \cite{pople1999nobel}. Given this target, the overall shot count becomes prohibitively large, even without considering measurements for gradient estimation. 

Currently, significant efforts focus on reducing the number of terms to be measured. A natural approach involves grouping mutually commuting terms, since they share a common eigenbasis, making expectation value estimation amenable to classical postprocessing. However, since quantum hardware measurements are typically constrained to the computational basis, the required rotation operations, which scale depth-wise as $\mathcal{O}(N^2/ log(N))$, are not implementable on current hardware \cite{verteletskyi2020measurement}. Consequently, many schemes have been proposed that consider more restrictive classes of commuting groupings, limiting the measurement basis to single-qubit measurements. A recent review can be found in \cite{patel2025quantum}.

Employing auxiliary qubits as in the Hadamard test (see Fig.~\ref{fig:hadamard}) is a possible avenue, but this approach is reserved for the fault-tolerant era because the routine involves deep circuits arising from numerous controlled-unitary operations. The shot count for estimating the overall expectation value of $H$ scales $\mathcal{O}(\lambda^2/\epsilon^2)$, where $\lambda$ is the 1-norm of $H$. Since resource requirements for \ac{QPE}, the workhorse for fault-tolerant quantum computing, also scale with the 1-norm, tremendous efforts have focused on developing decompositions to minimize this scaling, most successfully demonstrated in \cite{oumarou2024accelerating, patel2025global}.

A more NISQ-friendly, low circuit-depth approach is provided by the classical shadow protocol  \cite{huang2020predicting}. Since this technique becomes a key aspect of Chap.~\ref{sub:shadowsandstats}, we reserve detailed discussion for later. However, we note that it offers a highly promising framework for obtaining a density matrix estimator from polynomially many random measurements of the prepared quantum state, which can then be used to compute expectation values for a specific class of observables accurately. 

\subsubsection{Treatment of hardware errors in VQE}

As previously stated, \ac{VQE} is envisioned to be applied without error correction. Inevitably, estimated expectation values will be corrupted by hardware noise. The prevalent proposal to overcome this issue is \textit{error mitigation} \cite{cai2023quantum}, an empirical method that learns recurrent hardware errors from measurement data so their effects can be compensated through classical post-processing. We mention \ac{ZNE} here due to its intuitive nature and popularity, and refer to Appendix~\ref{appendix:cdr}, which is the scheme used in \cite{schroeder2024deterministic}. 
As the name suggests, \ac{ZNE} infers the noise-free estimator for the targeted inaccessible expectation value $\Tr(O \rho_0)$ by extrapolating data from expectation values with amplified noise. Specifically, the estimation for the noisy implementation of $\rho$ is given by $\Tr(O \rho_{\lambda_1})$. Noise is amplified by implementing even noisier versions through effectively inserting identity operations via applying unitaries and their inverses in succession, which makes the circuit deeper while implementing the same unitary. We then obtain $d$ estimators of the expectation values $\{ \Tr(O \rho_{\lambda_i}) \}_{i=1}^d$ with $d$ varying noise levels. References ~\cite{temme2017error, cai2023quantum}  showed that for small noise levels $\lambda$, the relationship between noise level and expectation value estimation follows a polynomial relation $\Tr (O \rho_\lambda) \approx \sum_{k = 0}^{d-1} \beta_k \frac{\lambda^k}{k!}$.  For the best-fit parameters $\beta_k$, the zero-noise expectation can be calculated at  $\lambda = 0$. 

A simple thought experiment illustrates that error mitigation always requires exponential sampling overhead to converge to the error-free output: Imagine access to an oracle that can differentiate between faulty and error-free outputs. Even with such a powerful oracle, given a noise rate of $\lambda$, only $e^{-\lambda}$ samples will be noise-free. To obtain sufficient noise-free samples, we need to collect at least $e^\lambda$ samples \cite{cai2023quantum}. Thus, error mitigation will always be only a workaround until error-correcting codes become experimentally available, only suitable for suppressing small error rates in prototypical implementations, but not feasible for elevating NISQ to the computational power of fault-tolerant quantum computers. Yet, experts claim that one should not be too pessimistic about the efficacy of error mitigation \cite{zimboras2025myths}. Since the sample overhead is exponential but the exponent scales linearly with the error rate $\epsilon$, qubit count $N$, and the circuit depth $L$ $\mathcal{O}(e^{\epsilon N L})$ \cite{tsubouchi2023universal}, in practice if $NL \propto \epsilon^{-1}$, error mitigation is feasible. Also, error mitigation will be important in the early fault-tolerant regime to correct for small residual errors \cite{dutkiewicz2024error} or to reduce hardware requirements of error correcting codes \cite{suzuki2022quantum}.

\subsubsection{Caveats of VQE}

Despite the widespread adoption of \acp{VQE} as the leading paradigm for quantum applications on NISQ devices \cite{peruzzo2014variational, kandala2017hardware, google2020hartree}, more and more theoretical and experimental evidence raises fundamental questions about their potential for demonstrating quantum advantage. This section examines the key limitations challenging the viability of VQE. 

In recent years, numerous approaches have rigorously demonstrated troubling aspects of VQE. The classical optimization of quantum ansätze has been shown to be NP-hard \cite{bittel2021training}, while training suffers from vanishing gradients, known as barren plateaus \cite{mcclean2018barren}, which emerge in universal ansätze. Paradoxically, evidence suggests that barren plateau-free ansätze are classically simulable \cite{cerezo2023does}, alongside other classical simulability results  \cite{bermejo2024quantum, angrisani2024classically}. Furthermore, serious concerns arise from sampling complexity. Even under ideal conditions, achieving chemical accuracy requires sampling $1/\epsilon^2 \approx 10^6$ times per term. Additionally, error mitigation resources scale exponentially even for logarithmically shallow circuit depths \cite{quek2022exponentially}. These factors collectively justify the healthy skepticism surrounding VQE as a vehicle for demonstrating quantum advantage.

\subsection{Synopsis of this thesis concerning VQE}

This thesis approaches \ac{VQE} from two perspectives: implementation and application. 

Regarding \textit{implementation}, while quantum computation is typically discussed with \ac{GBQC} in mind, there exists an equivalent but different formulation of quantum computing called \ac{MBQC} \cite{raussendorf2003measurement}. This approach implements quantum computing through quantum gate teleportation on large pre-prepared entangled resource states rather than sequential gate applications on static qubits. This paradigm offers an inherent gate-depth to qubit-count trade-off compared to \ac{GBQC}. Our work presented in \cite{schroeder2024deterministic} builds upon the primary proposal of \ac{MBVQE} \cite{ferguson2021measurement} and is replicated in Chapter~\ref{chap:mbvqe}.

The second perspective concerns \textit{application}. Given the aforementioned challenges, it appears infeasible to expect \ac{VQE} to become practically viable for directly estimating ground state energies in the near term. However, ideas have emerged to use \ac{VQE} to generate quantumly efficient but classically intractable trial wave functions for classical workflows. Specifically, a quantum-enhanced auxiliary-field quantum Monte Carlo method has been proposed in Ref.~\cite{huggins2022unbiasing}. Within \ac{QUTAC}, the author investigated the efficacy and applicability of this approach on industrially motivated test systems, resulting in two publications. These results are summarized in Chapter~\ref{chap:chem}, together with a brief introduction to quantum chemistry in the quantum computing context. Additionally, a more ``classical'' use case for \ac{VQE} has been explored in \ac{NLP} through an intra-university collaboration, yielding one publication that will be discussed with a focus on classical data encoding in Chapter~\ref{chap:qnlp}.

\end{document}